---
title: "RNA-Seq QC and DESeq2 Analysis"
author: "SK"
date: "3/8/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

Load any packages you will need for the RMD in the first R chunk -- that way you can load them all at once by running the whole chunk 

```{r load libraries, warning=FALSE}
suppressPackageStartupMessages({
library(dplyr)
library(tidyr)
library(ggplot2)
library(calibrate)
library(tibble)
library(enrichplot)
library(knitr)
library(kableExtra)
library(reshape2)
library(pheatmap)
library(gprofiler2)
library(vidger)
library(patchwork)
library(ggVennDiagram)
library(edgeR)
library(hypeR)
  library(DESeq2)
  library(RColorBrewer)
  library(ggrepel)
  library(EnhancedVolcano)
  library(data.table)
  library(geneplotter)
})

```

Load in your counts data -- should be in the form of a counts matrix
```{r load data, cache=TRUE}
# Enter file path to your counts matrix. You may need to use a different function to read in your data -- it depends on the format you saved it as.
wd = "/Users/sarafiller/Desktop/Thermofly/raw_rRNA"
setwd(wd)
counts<- read.csv("rRNA_depletion_counts.csv")


colnames(counts)
row.names(counts)<- counts$X
counts<-counts[,-1] # remove first column because genes now assigned to row.names
counts<-counts[,-7] # remove the 4th replicate of cold10

# Fill in with your main treatment groups using the rep() and factor() functions

condition <- factor(c(rep("Control25", 3),  rep("Cold10",3), rep("Hot34", 3), rep("Hot37",3), rep("Cold4",3)))
str(counts)
# You can add more meta data in another vector to use for DESeq analysis, but we will keep it simple. For example you could write: species <- factor(c(rep("D. melanogaster", 8), rep("D. pseudoobscura", 8)))

coldata <- data.frame(row.names=colnames(counts), condition)

# Create DESeq2Dataset object
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = coldata,
                              design= ~ condition)

# keep only rows where row sums are >= 10 
keep <- rowSums(counts(dds)) >= 10
dds<- dds[keep,]

# Run DESeq2 analysis/ make dds object 
dds<- DESeq(dds)
colnames(dds) <- colnames(counts)

```

Step 2: DEG analysis using DESeq2
## Run contrasts to obtain DEG results
```{r Run contrasts, warning= F, echo=FALSE, cache = T}
##### DESeq2 Contrasts ########
# Set up a new pairwise contrast for each of your conditions -- repeat these 7 lines, changing object names to store DEG lists in uniquely named objects. You can also adjust the "signficance" threshold by altering the last line.

# Contrast example: Hot 37º vs Control 25º
res_hot37v25 <- results(dds, contrast=c('condition','Hot37','Control25')) 
dim(res_hot37v25)
res_hot37v25 <- res_hot37v25[complete.cases(res_hot37v25),]  #remove any rows with NA
dim(res_hot37v25)
#rank genes by ordering them by p adjusted value
result_hot37v25 <- res_hot37v25[order(res_hot37v25$padj), ]
head(result_hot37v25)
# extract DEGs
#change to a dataframe, subset out genes with an abs value logfold change   from the control group greater than 1 (negative > less expression, positive > more expression)
#only include genes with p<0.05
result_hot37v25DEG<-
data.frame(result_hot37v25[which(abs(result_hot37v25$log2FoldChange) > 1 & result_hot37v25$padj < 0.05),])
# result_hot37v25DEG is now a dataframe with your DEGs
#upregualted DEGs
upreg_hot37v25DEG<-
  data.frame(result_hot37v25DEG[which(result_hot37v25DEG$log2FoldChange > 1),])
#extract downregulated DEGs
downreg_hot37v25DEG<-
  data.frame(result_hot37v25DEG[which(result_hot37v25DEG$log2FoldChange < 1),])


# Contrast example: Hot 34º vs Control 25º
res_hot34v25 <- results(dds, contrast=c('condition','Hot34','Control25')) 
dim(res_hot34v25)
res_hot34v25 <- res_hot34v25[complete.cases(res_hot34v25),]  #remove any rows with NA
dim(res_hot34v25)
#order by p adjusted value
result_hot34v25 <- res_hot34v25[order(res_hot34v25$padj), ]
head(result_hot34v25)
# extract DEGs
result_hot34v25DEG<-
data.frame(result_hot34v25[which(abs(result_hot34v25$log2FoldChange) > 1 & result_hot34v25$padj < 0.05),])
#DEGs = the number rows in the dataframe, nrow(result_hot34v25DEG)
#upregualted DEGs
upreg_hot34v25DEG<-
  data.frame(result_hot34v25DEG[which(result_hot34v25DEG$log2FoldChange > 1),])
#extract downregulated DEGs
downreg_hot34v25DEG<-
  data.frame(result_hot34v25DEG[which(result_hot34v25DEG$log2FoldChange < 1),])

# Contrast example: Cold 10º vs Control 25º
res_cold10v25 <- results(dds, contrast=c('condition','Cold10','Control25')) 
dim(res_cold10v25)
res_cold10v25 <- res_cold10v25[complete.cases(res_cold10v25),]  #remove any rows with NA
dim(res_cold10v25)
#order by p adjusted value
result_cold10v25 <- res_cold10v25[order(res_cold10v25$padj), ]
head(result_cold10v25)
# extract DEGs
result_cold10v25DEG<-
data.frame(result_cold10v25[which(abs(result_cold10v25$log2FoldChange) > 1 & result_cold10v25$padj < 0.05),])
#upregualted DEGs
upreg_cold10v25DEG<-
  data.frame(result_cold10v25DEG[which(result_cold10v25DEG$log2FoldChange > 1),])
#extract downregulated DEGs
downreg_cold10v25DEG<-
  data.frame(result_cold10v25DEG[which(result_cold10v25DEG$log2FoldChange < 1),])

# Contrast example: Cold 4º vs Control 25º
res_cold4v25 <- results(dds, contrast=c('condition','Cold4','Control25'))
dim(res_cold4v25)
res_cold4v25 <- res_cold4v25[complete.cases(res_cold4v25),]  #remove any rows with NA
dim(res_cold4v25)
#order by p adjusted value
result_cold4v25 <- res_cold4v25[order(res_cold4v25$padj), ]
head(result_cold4v25)
# extract DEGs
result_cold4v25DEG<-
data.frame(result_cold4v25[which(abs(result_cold4v25$log2FoldChange) > 1 & result_cold4v25$padj < 0.05),])

# extract upregulatedDEGs
upreg_cold4v25DEG<-
  data.frame(result_cold4v25DEG[which(result_cold4v25DEG$log2FoldChange > 1),])
#extract downregulated DEGs
downreg_cold4v25DEG<-
  data.frame(result_cold4v25DEG[which(result_cold4v25DEG$log2FoldChange < 1),])
```
# Make a dataframe called DEGdata with three cols, treatment,variable: up/down, number
```{r}
list1<- list("result_cold4v25DEG"= result_cold4v25DEG, "result_cold10v25DEG"= result_cold10v25DEG, "result_hot34v25DEG"=result_hot34v25DEG, "result_hot37v25DEG"= result_hot37v25DEG)

deg_data<- data.frame("treat"= c(names(list1)), "up" = NA, "down"= NA)
deg_data$treat<- gsub("result_", "", deg_data$treat)
deg_data$treat<- gsub("v25DEG", "", deg_data$treat)

for(i in 1:4){
  deg_data$up[i]<- length(which(list1[[i]][["log2FoldChange"]] >= 1))
    deg_data$down[i]<- length(which(list1[[i]][["log2FoldChange"]] <= -1))
}

deg_data<- melt(deg_data)
deg_data$treat <- factor( deg_data$treat, levels= c("cold4", "cold10", "hot34", "hot37"))

p<- ggplot(deg_data, aes(x= treat, y= value, fill= variable)) + geom_bar(position=position_dodge(), aes(y=value), stat="identity") +
  geom_text(aes(label=value), vjust=-0.25, color="black", position = position_dodge(0.9), size=3.5) + theme_bw()  + scale_y_continuous(limits = c(0, 1790), expand = c(0,0)) + ggtitle("DEGs Relative to 25ºC Control") + scale_fill_manual(values= c("red", "lightblue"))
# + scale_fill_manual(values=colors))
p
#View(p) to show the plot
```

## VOLCANO PLOTS
## Highlight DEGs in volcano plots
```{r Enhanced Volcano plots, warning=FALSE, echo=FALSE, cache= T} 
# Repeat for each condition (must highlight and run each separately)
# Hot 37º vs Control 25º
res_37v25 <- lfcShrink(dds, contrast=c('condition','Hot37','Control25'), type = "normal") 

data<- res_37v25
title<- "Hot 37º vs Control 25º"

# Hot 34º vs Control 25º
res_34v25 <- lfcShrink(dds, contrast=c('condition','Hot34','Control25'), type = "normal") 

data<- res_34v25
title<- "Hot 34º vs Control 25º"

# Cold 10º vs Control 25º
res_10v25 <- lfcShrink(dds, contrast=c('condition','Cold10','Control25'), type = "normal") 

data<- res_10v25
title<- "Cold 10º vs Control 25º"

# Cold 4º vs Control 25º
res_4v25 <- lfcShrink(dds, contrast=c('condition','Cold4','Control25'), type = "normal") 

data<- res_4v25
title<- "Cold 4º vs Control 25º"

###############################

FC <- 1
p <- 5e-2

keyvals <- rep('grey50', nrow(data))
names(keyvals) <- rep('NS', nrow(data))

keyvals[which(abs(data$log2FoldChange) > FC & data$padj > p)] <- 'grey50'
names(keyvals)[which(abs(data$log2FoldChange) > FC & data$padj > p)] <- 'log2FoldChange'

keyvals[which(abs(data$log2FoldChange) < FC & data$padj < p)] <- 'grey50'
names(keyvals)[which(abs(data$log2FoldChange)  < FC & data$padj < p)] <- '-Log10Q'

keyvals[which(data$log2FoldChange < -FC & data$padj < p)] <- 'red2'
names(keyvals)[which(data$log2FoldChange  < -FC & data$padj < p)] <- 'Signif. down-regulated'

keyvals[which(data$log2FoldChange > FC & data$padj < p)] <- 'red2'
names(keyvals)[which(data$log2FoldChange > FC & data$padj < p)] <- 'Signif. up-regulated'

unique(keyvals)
unique(names(keyvals))

# plot
volcano <- EnhancedVolcano(data,
                           lab = rownames(data),
                           x = 'log2FoldChange',
                           y = 'pvalue',
                           selectLab = NULL,
                           ylim = c(0,20),
                           xlim = c(-3,3),
                           xlab = bquote(~Log[2]~ 'fold change'),
                           ylab = bquote(~-Log[10] ~ italic(P)),
                           title = paste0(title),
                           pCutoff = 5e-2,
                           FCcutoff = 1,
                           pointSize = 1,
                           labSize = 3.0,
                           labCol = 'black',
                           labFace = 'bold',
                           boxedLabels = F,
                           #shape = c(6, 4, 2, 11, 15),
                           colCustom = keyvals,
                           colAlpha = 1,
                           legendPosition = 'right',
                           legendLabSize = 5,
                           legendIconSize = 5.0,
                           drawConnectors = FALSE,
                           widthConnectors = 0.5,
                           colConnectors = 'grey50',
                           gridlines.major = FALSE,
                           gridlines.minor = FALSE,
                           border = 'partial',
                           borderWidth = 0.5,
                           borderColour = 'black')

volcano


# Change axes, colors and labeling within the plotting function
```

## Make a histogram of your DEGs
```{r Histogram of up and down DEGs at each temperature}
# use ggplot to make a histogram of your DEGs -- there are many ways to do this!
#y is counts, x is based on the log2 fold change if it >1 or <1

# My simple bar plot for one comparison

bar_plot_df<-result_hot37v25DEG%>%
  summarise(up_reg=log2FoldChange>1,
            down_reg=log2FoldChange<1)
#%>%
  #group_by(up_reg)%>%summarize(count=n())  #this distills it to just the counts

bar_plot_df%>%
  ggplot(aes(up_reg))+
  geom_bar()+scale_x_discrete(labels=c("TRUE"="up_reg","FALSE"="down_reg"),name="Number DEGs up or down regulated")

# Sophie's Data (from slack for bar graphs with all data altogether) ----
# repeat 2 plots above

```

## Clustered heatmap of DEG expression
```{r clustered heatmap of DEGs, warning=FALSE, echo= F, cache= T}
# look at expression patterns of genes that cluster in the different conditions - favor the DEGs only


# Concatenate all the unique DEGs identified in each of the contrasts -- gene names are stored in the result_* objects in the row names
uniquedeg<- unique(c(row.names(result_cold4v25DEG), row.names(result_cold10v25DEG), row.names(result_hot34v25DEG), row.names(result_hot37v25DEG)))

# Write list of DEGs to a file
write.table(uniquedeg,"rRNA_DEGs.csv", row.names=FALSE,col.names=FALSE)

# normalize data (uses edgeR)
y <- DGEList(counts=counts,group=condition)
y <- calcNormFactors(y)

lcpm <- cpm(y, log=TRUE) #log cpm
# subset the unique degs
filtered <- lcpm[intersect(row.names(lcpm), as.character(uniquedeg)),]

#add annotations
condition <- factor(c(rep("Cold10", 3),  rep("Control25",3), rep("Hot34", 3), rep("Hot37",3), rep("Cold4",3)))
anno = data.frame("category" = condition)

rownames(anno) = colnames(filtered)

# plot heatmap with kmeans to get clusters
out <- pheatmap(filtered, 
          cluster_rows = T, 
          cluster_cols = T, 
          clustering_method = "mcquitty", 
          clustering_distance_rows = "euclidean",  
          show_rownames = F, 
          annotation_col = anno, 
          scale = "row", 
          show_colnames = F, 
          cutree_rows = 3,
          color= colorRampPalette(rev(brewer.pal(n = 7, name ="RdBu")))(100),
          clustering_distance_cols = "correlation",
          annotation_names_col = F,
          angle_col = "45", 
          kmeans_k = 3 # need to break means into clusters
          )

clusters <- cutree(out$tree_row, k=3)[out$tree_row[["order"]]]

annot_row <- data.frame(row.names = names(clusters),
                        cluster = as.factor(clusters))
ann_color= list(category= c(Cold4= "#357EBDFF", Cold10= "#46B8DAFF", Control25= "#5CB85CFF", Hot34= "#EEA236FF", Hot37= "#D43F3AFF" ))

pheatmap(filtered, 
          cluster_rows = T, 
          cluster_cols = T, 
          clustering_method = "mcquitty", 
          clustering_distance_rows = "euclidean",  
          show_rownames = F, 
          annotation_col = anno, 
          scale = "row",
          show_colnames = F,
          cutree_rows = 3,
          color= colorRampPalette(rev(brewer.pal(n = 7, name ="RdBu")))(100),
          clustering_distance_cols = "correlation",
          annotation_names_col = F,
          angle_col = "45",
        # annotation_row = annot_row, 
         main = "Expression of DEGs",
         annotation_colors = ann_color[1]
          )

geneclusters <- out[["kmeans"]][["cluster"]]

# Get cluster genes from the heatmap -- note the number of clusters will vary

# get cluster1 genes
cl1 <- geneclusters[geneclusters==1]
cl1 <- as.data.frame(cl1)
cluster1 <- row.names(cl1)
head(cluster1)
length(cluster1)

#get cluster2 genes
cl2 <- geneclusters[geneclusters==2]
cl2 <- as.data.frame(cl2)
cluster2 <- row.names(cl2)
head(cluster2)
length(cluster2)

#get cluster3 genes
cl3 <- geneclusters[geneclusters==3]
cl3 <- as.data.frame(cl3)
cluster3 <- row.names(cl3)
head(cluster3)
length(cluster3)

#get cluster4 genes
cl4 <- geneclusters[geneclusters==4]
cl4 <- as.data.frame(cl4)
cluster4 <- row.names(cl4)
head(cluster4)
length(cluster4)

#get cluster5 genes
cl5 <- geneclusters[geneclusters==5]
cl5 <- as.data.frame(cl5)
cluster5 <- row.names(cl5)
head(cluster5)
length(cluster5)

#make dataframe of all cluster genes to easily pull a cluster for pathway
cluster_df <- as.data.frame(geneclusters)
#write.csv(cluster_df, "fly_geneclusters.csv", quote= F)
```

## Create a csv of a table containing all degs by condition
```{r}
# make a new df for each condition containing all DEGs and add a col saying 
#   from which comparison they came from 
comparison_list<- list("cold4v25"= result_cold4v25DEG, 
             "cold10v25"= result_cold10v25DEG, 
             "hot34v25"=result_hot34v25DEG, 
             "hot37v25"= result_hot37v25DEG)

for (i in names(comparison_list)){
  print(i)
  deseq_obj=paste("result_",i,"DEG", sep="")
  df_name=(paste(i,"_df",sep=""))
  df<-comparison_list[[i]] # make a new df from one of the DEG dfs in comparison_list
  df$comparison<-c(i) # add a new column telling you the name of the comparison
  assign(df_name,df) #assigns the corresponding comparison name to the df
  # would be nice to add each newly made df to a list of dfs and 
  # then i could just bind all the dfs in the list
}

# For each condition, then group all rows together into a larger df
degs_by_condition<-rbind(cold4v25_df,cold10v25_df,hot34v25_df,hot37v25_df)

# Write to a csv
write.table(degs_by_condition,"rRNA_degs_by_condtion.csv", sep=',',
            quote=FALSE)
```



